Data Management and Domain Adjusting Strategies for Implementing Parallel SPH Method in JPUE Simulation

### 2016-01-08 Paper for submission.tex
minor modification

In both SPHysics and DualSPHysics\cite{crespo2015dualsphysics}, the physical quantities of each particle (position, velocity, density…) are stored in arrays, and the particles (and the arrays with particle data) are reordered following the order of the cells. This has two advantages: 1) access pattern is more regular and more efficient, 2) it is ease to identify the particles that belong to a cell by using a range since the first particle of each cell is known. But adding, deleting and especially accessing of particles are cumbersome. Ferrari\cite{ferrari2009new} adopted linked lists using pointers so that particles can be deleted or added during the simulation. Storage problems caused by fix-size arrays are thereby also eliminated. We define C++ classes which contains all data of particle and bucket. As for the management of data, we adopt hash tables to store pointers to particles and buckets, which give us not only flexibility of deleting and adding element, but also quicker access compared with linked list. Instead of using the "nature manner" to number particles, we adopt SFC based index to give each particle and background bucket an unique identifier -- a strategy known to preserve some locality at minimal cost. The SFC based numbering strategy is further extended to include time step information so that particles added at the same position but different time will have different identifiers. 

@@

SFC-based indexing scheme can guarantee uniqueness of particle identifier only in simple scenarios when particles are added once while setting up initial condition. In some situations, new particles need to be added while simulation. For example, new particles need to be added at the bottom of the eject vent for JPUE simulation (see figure \ref{fig:Particle_adding_with_link}). To distinguish particles added at the "same place" (the small area, all points within which will be mapped to the same $k$.) at different time steps, we extend the SFC-based key to time-dependent SFC based key by including date of birth of particles into the key. Then the time-dependent SFC based key can be written as: $[k,t]$, where $t$ is the time step. The map $h_n$ will become:

@@

The most basic data structure of SPH are particle, for problem description, and bucket, for neighbors searches and domain decomposition. Both are defined as classes in C++. Information that is contained in particle class can be categorized into six categories: ID(the key), affiliate(rank of the process that the particle belongs to), primitive variables (variables show up as unknowns in governing equations, e.g. density, velocity), secondary variables (physical properties that can be computed from primitive variables, they are stored to avoid repeated computations eg. pressure, temperature), flags (indicators, such as indicator for ghost particle and real particles, indicator for particles of different phases) and neighbor information (it is a vector of particle keys in our application). Similarly,  Information that is contained in  a bucket class can also be categorized into different categories: ID, affiliate, domain information (maximum and minimum coordinates, boundary information), flags (indicators, such as indicator for guest and non-guest, indicator for active and inactive), neighbor information (keys of 27 neighbor buckets including its own key) and owned particles (it is a vector of particle keys in our application). Objects defined based on these two classes are then accessed through hash tables.

@@

Figure \ref{fig:SFC_particles_buckets_partition} shows how domain is decomposed based on partition of SFC of buckets. The particles are automatically split into several groups along with buckets that contain them. As SFC of buckets is a curve in the three dimensional space, partition of this curve will automatically lead to 3D domain decomposition. A 2D domain decomposition based on SFC of footprint buckets projected by three dimensional buckets was adopted by Kumar et al \cite{kumar2013parallel}. A comparison of the scalability of these two schemes (see figure \ref{fig:2D_vs_3D_efficiency}) confirms that 3D domain decomposition is a better choice when processes number is larger. 

@@

\caption{Execution time of test case 1 and test case 2}

@@

The initial domain is $[-4800m,4800m]\times [-4800m,4800m] \times [0m, 6000m]$, with smoothing length (we set initial intervals between particles equal to smoothing length) equals to 200m and 100m respectively for test case1 and test case2, respectively. The computational work load of test case 2 is 8 times of that of the test case 1. The simulations run for 20s physical time.  Wall clock time and speed up are shown in figure \ref{fig:2cases_time} and \ref{fig:2cases_efficiency}. Linear speed up is observed when number of processes is smaller than 16. Test case2 shows better speed up than test case1 which implies that the overhead of strong scalability can be increased by increasing total amount of work load. We also compared the performance of 2D domain decomposition and 3D domain decomposition. As shown by figure \ref{fig:2D_vs_3D_efficiency}, 2D domain decomposition shows a little bit better speed up than 3D domain decomposition for 32 processes. This can be explained by the fact that 2D domain decomposition can get more regular subdomain and as a results, need less communication. When number of processes increase, it becomes harder for 2D decomposition to balance workload, that is why 3D domain decomposition has better speed up for processes number equals to 64 and 128. The weak scalability test is conducted with the same initial domain and various smoothing length. Each simulation runs for 400 loops. The average number of real particles of each process keep constant at $25900$. As shown in figure \ref{fig:weak_loglog}, simulation time are almost constant with some minor fluctutations around the average value. These fluctuations are due to sharing nodes with other users. In addition, difference in simulation time for different processors are obvious.

@@

A typical simulation result of volcano plume is shown in figure \ref{fig:Plume}. Input parameters for this simulation are exactly the same as run P1 in reference \cite{suzuki2005numerical}.

>>>>

In both SPHysics and DualSPHysics\cite{crespo2015dualsphysics}, the physical quantities of each particle (position, velocity, density...) are stored in arrays, and the particles (and the arrays with particle data) are reordered following the order of the cells. This has two advantages: 1) access pattern is more regular and more efficient, 2) it is ease to identify the particles that belong to a cell by using a range since the first particle of each cell is known. But adding, deleting and especially accessing of particles are cumbersome. Ferrari\cite{ferrari2009new} adopted linked lists using pointers so that particles can be deleted or added during the simulation. Storage problems caused by fix-size arrays are thereby also eliminated. We define C++ classes which contain all data of particle and bucket. As for the management of data, we adopt hash tables to store pointers to particles and buckets, which give us not only flexibility of deleting and adding element, but also quicker access compared with linked list. Instead of using the "nature manner" to number particles, we adopt SFC based index to give each particle and background bucket an unique identifier -- a strategy known to preserve some locality at minimal cost. The SFC based numbering strategy is further extended to include time step information so that particles added at the same position but different time will have different identifiers. 

@@

SFC-based indexing scheme can guarantee uniqueness of particle identifier only in simple scenarios when particles are added once while setting up initial condition. In some situations, new particles need to be added while simulation. For example, new particles need to be added at the bottom of the eject vent for JPUE simulation (see figure \ref{fig:Particle_adding_with_link}). To distinguish particles added at the "same place" (the small area, all points within which will be mapped to the same $k$.) at different time steps, we extend the SFC-based key to time-dependent SFC based key by including date of birth of particles into their keys. Then the time-dependent SFC based key can be written as: $[k,t]$, where $t$ is the time step. The map $h_n$ will become:

@@

The most basic data structure of SPH are particle, for problem description, and bucket, for neighbors searches and domain decomposition. Both are defined as classes in C++. Information that is contained in particle class can be categorized into six categories: ID(the key), affiliate(rank of the process that the particle belongs to), primitive variables (variables show up as unknowns in governing equations, e.g. density, velocity), secondary variables (physical properties that can be computed from primitive variables, they are stored to avoid repeated computations eg. pressure, temperature), flags (indicators, such as indicator for ghost particle and real particles, indicator for particles of different phases) and neighbor information (it is a vector of particle keys in our application). Similarly,  Information that is contained in  a bucket class can also be categorized into different categories: ID, affiliate, dimension information (maximum and minimum coordinates, boundary information), flags (indicators, such as indicator for guest and non-guest, indicator for active and inactive), neighbor information (keys of 27 neighbor buckets including its own key) and owned particles (it is a vector of particle keys in our application). Objects defined based on these two classes are then accessed through hash tables.

@@

Figure \ref{fig:SFC_domain_decomposition} shows how domain is decomposed based on partition of SFC of buckets. The particles are automatically split into several groups along with buckets that contain them. As SFC of buckets is a curve in the three dimensional space, partition of this curve will automatically lead to 3D domain decomposition. A 2D domain decomposition based on SFC of footprint buckets projected by three dimensional buckets was adopted by Kumar et al \cite{kumar2013parallel}. A comparison of the scalability of these two schemes (see figure \ref{fig:2D_vs_3D_efficiency}) confirms that 3D domain decomposition is a better choice when processes number is larger. 

@@

\caption{Wall clock time of test case1 and test case2}

@@

The initial domain is $[-4.8km,4.8km]\times [-4.8km,4.8km] \times [0km, 6km]$, with smoothing length (we set initial intervals between particles equal to smoothing length) equals to 200m and 100m respectively for test case1 and test case2, respectively. The computational work load of test case 2 is 8 times of that of the test case 1. The simulations run for 20s physical time.  Wall clock time and speed up are shown in figure \ref{fig:2cases_time} and \ref{fig:2cases_efficiency}. Linear speed up is observed when number of processes is smaller than 16. Test case2 shows better speed up than test case1 which implies that the overhead of strong scalability can be increased by increasing total amount of work load. We also compared the performance of 2D domain decomposition and 3D domain decomposition. As shown by figure \ref{fig:2D_vs_3D_efficiency}, 2D domain decomposition shows a little bit better speed up than 3D domain decomposition for 32 processes. This can be explained by the fact that 2D domain decomposition can get more regular subdomain and as a results, need less communication. When number of processes increase, it becomes harder for 2D decomposition to balance workload, that is why 3D domain decomposition has better speed up for processes number equals to 64 and 128. The weak scalability test is conducted with the same initial domain and various smoothing length. Each simulation runs for 400 loops. The average number of real particles of each process keep constant at $25900$. As shown in figure \ref{fig:weak_loglog}, simulation time are almost constant with some minor fluctutations around the average value. These fluctuations are due to sharing nodes with other users. In addition, difference in simulation time for different processors are obvious.\\

@@

A typical simulation result of volcano plume is shown in figure \ref{fig:Plume}. Input parameters for this simulation are exactly the same as "Run P1" in reference \cite{suzuki2005numerical}.
